% R bootcamp, Module 1: Basics
% August 2013, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
```

# R as a calculator

Note: Remember to start recording.

```{r computing,echo=TRUE,results='markup'}
2 + 2 # add numbers
2 * pi # multiply by a constant
7 + runif(1) # add a random number
3^4 # powers
sqrt(4^4) # functions
log(10)
log(100, base = 10)
23 %/% 2 
23 %% 2

# scientific notation
5000000000 * 1000
5e9 * 1e3
```

Think of a mathematical operation you need - can you guess how to do it in R?

# Assigning values to R objects

```{sh}
ls
```

A key action in R is to store values in the form of R objects, and to examine the value of R objects.

```{r}
val <- 3
val
print(val)

Val <- 7 # case-sensitive!
print(c(val, Val))
```

We can work with (and store) sequences and repetitions
```{r}
mySeq <- 1:6
mySeq

myOtherSeq <- seq(1.1, 11.1, by = 2)
myOtherSeq
length(myOtherSeq)

fours <- rep(4, 6)
fours

# This is a comment: here is an example of non-numeric data
depts <- c('espm', 'pmb', 'stats')
depts
```

If we don't assign the output of a command to an object, we haven't saved it for later use.

R gives us a lot of flexibility (within certain rules) for assigning to (parts of) objects from (parts of) other objects.

# How to be [lazy](http://dilbert.com/strips/comic/2005-05-29/)

If you're starting to type something you've typed before, or the long name of an R object or function, STOP!  You likely don't need to type all of that.

- Tab completion
- Command history 
    * up/down arrows
    * Ctrl-{up arrow} or Command-{up arrow}
- RStudio: select a line or block for execution
- Put your code in a file and use `source()`. For example: `source('myRcodeFile.R')`

Are there other tricks that anyone knows of?

# Vectors in R

The most basic form of an R object is a vector. In fact, individual (scalar) values are vectors of length one. 

We can concatenate values into a vector with `c()`.

```{r}
# numeric vector
nums <- c(1.1, 3, -5.7)
devs <- rnorm(5)
devs

# integer vector
ints <- c(1L, 5L, -3L) # force storage as integer not decimal number
# 'L' is for 'long integer' (historical)

idevs <- sample(ints, 100, replace = TRUE)

# character vector
chars <- c('hi', 'hallo', "mother's", 'father\'s', 
   "She said, 'hi'", "He said, \"hi\"" )
chars
cat(chars, sep = "\n")

# logical vector
bools <- c(TRUE, FALSE, TRUE)
bools
```

# Working with indices and subsets

```{r}
vals <- seq(2, 12, by = 2)
vals

vals[3]
vals[3:5]
vals[c(1, 3, 6)]
vals[-c(1, 3, 6)]
vals[c(rep(TRUE, 3), rep(FALSE, 2), TRUE)]
```
We can substitute values into vectors
```{r}
vals[4] <- -35
vals[1:2] <- 0

# How does R process these next subset operations?
vals <- rnorm(100)
vals[vals < 0] <- 0
vals[1:8]

set.seed(0)  # make random number generation repeatable
vals <- rnorm(100)
extremes <- vals[vals > 3]
extremes
# what happened?
```

# Vectorized calculations and comparisons

At the core of R is the idea of doing calculations on entire vectors.

```{r}
vec1 <- sample(1:5, 10, replace = TRUE)
vec2 <- sample(1:5, 10, replace = TRUE)
vec1
vec2
vec1 + vec2
vec1^vec2

vec1 >= vec2
vec1 <= 3

vec1 == vec2
vec1 != vec2
# careful: 
vec1 = vec2
identical(vec1, vec2)
```

An important related concept is that of recycling
```{r}
vec3 <- sample(1:5, 5, replace = TRUE)
vec4 <- sample(1:5, 3, replace = TRUE)
vec1
vec3
vec4 

vec1 + vec3
vec1 + vec4
```

Tell me what's going on. What choices were made by the R developers?

# R is a functional language

* Operations are carried out with functions. Functions take objects as inputs and return objects as outputs. 
* An analysis can be considered a pipeline of function calls, with output from a function used later in a subsequent operation as input to another function.
* Functions themselves are objects: 
```{r}
sum
class(sum)
```
* We can embed function calls: 
```{r}
hist(rnorm(1000))
```

# Getting help about a function

To get information about a function you know exists, use `help` or `?`, e.g., `?lm`. For information on a general topic, use `apropos` or `??`

```{r eval = FALSE}
help(lm)
?lm

?log

# ??'regr' # take out the '#' - there's an issue with Markdown processing...
```

# Basic kinds of R objects

Vectors are not the only kinds of R objects.

* Vectors: vectors of various types (numeric (i.e., decimal/floating point/double), integer, boolean, character), all items must be of the same type

* Matrices: matrices of various types, all items must be of the same type
```{r}
mat <- matrix(rnorm(9), nrow = 3)
t(mat) %*% mat
dim(mat)
``` 

* Lists: collections of disparate or complicated objects
```{r}
myList <- list(stuff = 3, mat = matrix(1:4, nrow = 2), 
   moreStuff = c("china", "japan"), list(5, "bear"))
myList
myList[[1]] # result is not (usually) a list (unless you have nested lists)
identical(myList[[1]], myList$stuff)
myList$moreStuff[2]
myList[[4]][[2]]
myList[1:3] # subset of a list is a list
myList$newOne <- 'more weird stuff'
names(myList)
```

Lists can be used as vectors of complicated objects. E.g., suppose you have a linear regression for each value of a stratifying variable. You could have a list of regression fits. Each regression fit will itself be a list, so you'll have a list of lists.

# Other classes of objects      

R has several approaches to object-oriented programming.  These are widely used, albeit a bit klunky. 

The most basic is 'S3' objects. These objects are generally built upon lists.

```{r}
y <- rnorm(10)
x <- rnorm(10)
mod <- lm(y ~ x)
class(mod)
is.list(mod)
names(mod)
mod$coefficients
mod[['coefficients']]
mod[[1]]
```

The magic of OOP here is that methods (i.e., functions) can be tailored to work specifically with specific kinds of objects.

```{r}
summary(y)
summary(mod)
```

What do you think R is doing behind the scenes?

Consider `summary.lm`.

# Converting between different types of objects

You can use the `as()` family of functions.

```{r}
ints <- 1:10
as.character(ints)
as.numeric(c('3.7', '4.8'))
```

Be careful: R tries to be helpful and convert between types/classes when it thinks it's a good idea. Sometimes it is overly optimistic. 

```{r}
indices <- c(1.7, 2.3)
ints[indices]
ints[0.999999999]
```



# Managing your objects

R has a number of functions for getting metadata about your objects. Some of this is built in to RStudio.

```{r}

vec1 <- 1:4
vec2 <- c(1, 2, 3, 4)

length(vec1)
str(vec1)
class(vec1)
typeof(vec1)
class(vec2)
typeof(vec2)
is.vector(vec1)
is.list(vec1)

is.list(myList)
is.vector(myList)

```

What have you learned? Does it make sense? 

# Managing and saving the workspace

R also has functions for learning about the collection of objects in your workspace. Some of this is built in to RStudio.

```{r}
ls()  # search the user workspace (global environment)
rm(x)    # delete a variable
ls()
ls.str() # list and describe variables
```

Finally we can save the objects in our R session:
```{r}
ls()
save.image('module1.Rda')
rm(list = ls())
ls()
load('module1.Rda') 
# the result of this may not be quite right in the slide version
ls()
```

Challenge: how would I find all of my objects that have 'x' in their names?

# A bit on plotting

R has several different plotting systems:

- *base* graphics
- *lattice* graphics
- *ggplot2* (an add-on package)

We'll see a little bit of *base* graphics here and then *lattice* and *ggplot2* tomorrow in Module 8.


